Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}

### 2014-08-25 interimreport2.tex
removed duplicate figures
Major revision of interim report
include data, figures, debugged code for mx variate analyses, etc.

As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002. A brief review of the history of measles in New Zealand was provided in our previous interim report. However, the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \today). In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.

@@

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks. In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection. The additional analyses included in this section are:
\begin{itemize}
\item Multivariate modelling to account for confounding within the univariate analyses.
\item Analyses of changing risk factors through time during outbreaks.
\end{itemize}

@@

Deviance Residuals: 
    Min       1Q   Median       3Q      Max  
-6.1174  -1.1562   0.0000   0.8599   6.7625

Coefficients:
                            Estimate Std. Error t value Pr(>|t|)    
(Intercept)               -8.268e+00  8.641e-01  -9.569 1.06e-08 ***
Age3-5                     6.931e-01  1.053e+00   0.658 0.518322    
Age6-17                    7.176e-01  9.064e-01   0.792 0.438300    
Age18-24                  -5.001e-01  1.149e+00  -0.435 0.668367    
Age25+                    -2.620e+00  1.158e+00  -2.262 0.035602 *  
EthnicityAsian             2.167e+00  1.203e+00   1.801 0.087576 .  
EthnicityMaori             1.772e+00  1.061e+00   1.671 0.111064    
EthnicityPacific           3.871e+00  9.405e-01   4.117 0.000587 ***
NZDep6-10                 -4.878e-01  2.608e-01  -1.870 0.076939 .  
Age3-5:EthnicityAsian     -2.431e-15  1.473e+00   0.000 1.000000    
Age6-17:EthnicityAsian    -2.473e+00  1.530e+00  -1.616 0.122649    
Age18-24:EthnicityAsian   -7.910e-01  1.622e+00  -0.488 0.631293    
Age25+:EthnicityAsian     -6.025e-01  1.792e+00  -0.336 0.740386    
Age3-5:EthnicityMaori     -2.845e+00  2.177e+00  -1.307 0.206941    
Age6-17:EthnicityMaori    -2.102e-01  1.121e+00  -0.187 0.853294    
Age18-24:EthnicityMaori   -8.881e-01  1.630e+00  -0.545 0.592102    
Age25+:EthnicityMaori     -1.090e+00  2.011e+00  -0.542 0.594088    
Age3-5:EthnicityPacific   -4.423e+00  2.581e+00  -1.714 0.102866    
Age6-17:EthnicityPacific  -2.253e+00  1.051e+00  -2.143 0.045264 *  
Age18-24:EthnicityPacific -8.321e-01  1.318e+00  -0.631 0.535272    
Age25+:EthnicityPacific   -2.718e-01  1.338e+00  -0.203 0.841210    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for quasipoisson family taken to be 16.26672)

    Null deviance: 3097.24  on 39  degrees of freedom
Residual deviance:  291.05  on 19  degrees of freedom
AIC: NA

Number of Fisher Scoring iterations: 6
\end{Soutput}
\end{Schunk}

@@

              Df Deviance Resid. Df Resid. Dev       F    Pr(>F)    
NULL                             39    3097.24                      
Age            4  1573.25        35    1523.98 24.1790 3.142e-07 ***
Ethnicity      3   738.20        32     785.78 15.1271 2.853e-05 ***
NZDep          1    55.39        31     730.39  3.4051   0.08064 .  
Age:Ethnicity 12   439.34        19     291.05  2.2507   0.05519 .  
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}
\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
     \caption{test}
     \label{fig:regn}
\end{figure}

@@

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{test}
     \label{fig:percap}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{test}
     \label{fig:percap}
\end{figure}

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{test}
     \label{fig:percap}
\end{figure}

\subsection{Modelling results and discussion}

@@

The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 7 19 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213. Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital. Only eight cases spent a week or more in the hospital. Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days. The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.

>>>>

As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002. A brief review of the history of measles in New Zealand was provided in our previous interim report. In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.

@@

A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks. In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection. The additional analyses included in this section are multivariate modelling to account for confounding within the univariate analyses.

\subsection{Risk analyses methods}
We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014. Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors. Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report. These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago. We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist.

The Otago denominator data provided were not to the same detail as the ESR case data. Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories. The denominator ethnicity was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided. After discussions with Otago we have provided results based on the best available data, though for smaller groups, some results may be unreliable as these are discussed below.

The prioritized ethnicities, therefore, are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African; Other Ethnicity; Residual Categories, though for this report only the first five are used, as these categories covered all the cases. Denominator data were classified according to the recommendations of Otago. Thus, with the 10 NZDep classes and the age classes above, this led us to have 250 categories. Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:agecases}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics (REF)).

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-002}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-004}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

Previous measles analyses, including two in New Zealand by Prof. Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. The overall population immunity rate is approximately 89\% and the breakdown of this by age category can be seen in Figure N.

\begin{figure}[h!]
\begin{center}
\begin{Schunk}
\begin{Soutput}
[1] NA
\end{Soutput}
\end{Schunk}
\includegraphics{interimreport2-006}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-008}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

This large number of categories lead to both zeroinflation and over-dispersion as there were many categories with zero cases in, particularly in the adult age classes. Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10. We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category. By doing so, we reduce the zeroinflation present in the data. Subsequent analyses also suggested that the Middle Eastern/Latin American/African category was over-represented (Figure~\ref{fig:percap}).

\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
     \caption{Cases per capita for Prioritized Ethnicities. Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity. Points are coloured by Ethnicity.}
     \label{fig:percap}
\end{figure}

However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, which have small population sizes, and which there were numberous issues with estimating the denominator data for. This category was also an outlier (Figure~\ref{fig:percap}), and thus reducing our ability to make inferences regarding the other classifications and risk factors. Thus, we removed the MELAA category for the below analyses. In the below analyses we used a quasipoission regression model to account for overdispersion in the data. We also account for differences in population sizes by using an offset term, the log(population size). Thus, we used a quasipoisson error structure and the following linear predictor:

\begin{equation} \label{eq:reg}
 log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _N (x_N) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
  \end{equation}

The numbers of cases per category can be seen in Table N.
\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
\hline
1-5&0-2&Asian&$ 11$\tabularnewline
6-10&0-2&Asian&$  8$\tabularnewline
1-5&3-5&Asian&$  1$\tabularnewline
1-5&6-17&Asian&$ 11$\tabularnewline
6-10&6-17&Asian&$  5$\tabularnewline
1-5&18-24&Asian&$  3$\tabularnewline
6-10&18-24&Asian&$  5$\tabularnewline
1-5&25+&Asian&$ 10$\tabularnewline
6-10&25+&Asian&$ 13$\tabularnewline
1-5&0-2&European&$ 83$\tabularnewline
6-10&0-2&European&$ 64$\tabularnewline
1-5&3-5&European&$ 42$\tabularnewline
6-10&3-5&European&$ 17$\tabularnewline
1-5&6-17&European&$219$\tabularnewline
6-10&6-17&European&$ 80$\tabularnewline
1-5&18-24&European&$ 34$\tabularnewline
6-10&18-24&European&$ 36$\tabularnewline
1-5&25+&European&$ 78$\tabularnewline
6-10&25+&European&$ 51$\tabularnewline
1-5&0-2&Maori&$ 18$\tabularnewline
6-10&0-2&Maori&$ 48$\tabularnewline
1-5&3-5&Maori&$  7$\tabularnewline
6-10&3-5&Maori&$ 11$\tabularnewline
1-5&6-17&Maori&$ 19$\tabularnewline
6-10&6-17&Maori&$ 92$\tabularnewline
1-5&18-24&Maori&$  5$\tabularnewline
6-10&18-24&Maori&$  8$\tabularnewline
1-5&25+&Maori&$  2$\tabularnewline
6-10&25+&Maori&$  6$\tabularnewline
6-10&0-2&MLA&$  3$\tabularnewline
6-10&3-5&MLA&$  1$\tabularnewline
6-10&6-17&MLA&$  1$\tabularnewline
6-10&18-24&MLA&$  2$\tabularnewline
6-10&25+&MLA&$  6$\tabularnewline
1-5&0-2&Pacific&$  5$\tabularnewline
6-10&0-2&Pacific&$ 58$\tabularnewline
6-10&3-5&Pacific&$  3$\tabularnewline
1-5&6-17&Pacific&$  5$\tabularnewline
6-10&6-17&Pacific&$ 22$\tabularnewline
1-5&18-24&Pacific&$  1$\tabularnewline
6-10&18-24&Pacific&$  8$\tabularnewline
1-5&25+&Pacific&$  2$\tabularnewline
6-10&25+&Pacific&$ 11$\tabularnewline
1-5&0-2&None&$  3$\tabularnewline
1-5&3-5&None&$  1$\tabularnewline
1-5&6-17&None&$  3$\tabularnewline
6-10&6-17&None&$  4$\tabularnewline
1-5&18-24&None&$  2$\tabularnewline
6-10&18-24&None&$  1$\tabularnewline
1-5&25+&None&$  5$\tabularnewline
6-10&25+&None&$  3$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:cases}
\end{table}

The numbers of cases per category can be seen in Table N.
\begin{table}\tiny
\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
\begin{center}
\begin{tabular}{lllrrr}
\hline\hline
\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Popn}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per Capita}\tabularnewline
\hline
1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
\hline
\end{tabular}\end{center}\label{table:percap}
\end{table}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-015}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-016}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-017}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-019}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-020}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\section{Regression analyses results}
\label{sub:regression_results}
The results are:

\begin{Schunk}
\begin{Sinput}
> summary(model2)
\end{Sinput}
\begin{Soutput}
Call:
glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
    Ethnicity:NZDep + offset(log(Popn)), family = "quasipoisson", 
    data = tp)

Deviance Residuals: 
     Min        1Q    Median        3Q       Max  
-1.98437  -0.58185  -0.00005   0.45759   1.47521

Coefficients:
                             Estimate Std. Error t value Pr(>|t|)    
(Intercept)                -6.422e+00  1.016e-01 -63.219  < 2e-16 ***
Age3-5                     -9.129e-01  1.794e-01  -5.088 5.62e-05 ***
Age6-17                    -7.653e-01  1.173e-01  -6.524 2.33e-06 ***
Age18-24                   -1.495e+00  1.692e-01  -8.836 2.43e-08 ***
Age25+                     -2.953e+00  1.405e-01 -21.024 4.16e-15 ***
EthnicityAsian              2.147e-03  3.167e-01   0.007  0.99466    
EthnicityMaori             -6.061e-02  2.267e-01  -0.267  0.79191    
EthnicityMLA               -1.820e+01  2.572e+03  -0.007  0.99442    
EthnicityPacific            6.950e-01  3.541e-01   1.963  0.06374 .  
NZDep6-10                  -3.339e-01  9.205e-02  -3.627  0.00168 ** 
Age3-5:EthnicityAsian      -2.032e+00  1.208e+00  -1.682  0.10813    
Age6-17:EthnicityAsian     -1.003e+00  4.125e-01  -2.432  0.02455 *  
Age18-24:EthnicityAsian    -8.417e-01  5.194e-01  -1.620  0.12079    
Age25+:EthnicityAsian       4.237e-01  3.873e-01   1.094  0.28707    
Age3-5:EthnicityMaori      -3.864e-01  3.578e-01  -1.080  0.29305    
Age6-17:EthnicityMaori     -7.940e-02  2.157e-01  -0.368  0.71660    
Age18-24:EthnicityMaori    -5.954e-01  3.917e-01  -1.520  0.14417    
Age25+:EthnicityMaori      -1.039e+00  4.579e-01  -2.270  0.03443 *  
Age3-5:EthnicityMLA        -1.857e-01  1.356e+00  -0.137  0.89245    
Age6-17:EthnicityMLA       -1.591e+00  1.349e+00  -1.179  0.25212    
Age18-24:EthnicityMLA       2.201e-01  1.076e+00   0.205  0.84001    
Age25+:EthnicityMLA         1.565e+00  8.351e-01   1.874  0.07556 .  
Age3-5:EthnicityPacific    -2.132e+00  7.110e-01  -2.998  0.00711 ** 
Age6-17:EthnicityPacific   -1.448e+00  2.924e-01  -4.953 7.66e-05 ***
Age18-24:EthnicityPacific  -9.854e-01  4.481e-01  -2.199  0.03979 *  
Age25+:EthnicityPacific    -5.958e-01  3.815e-01  -1.562  0.13406    
EthnicityAsian:NZDep6-10    1.330e-01  3.006e-01   0.442  0.66291    
EthnicityMaori:NZDep6-10    4.296e-01  2.080e-01   2.065  0.05210 .  
EthnicityMLA:NZDep6-10      1.890e+01  2.572e+03   0.007  0.99421    
EthnicityPacific:NZDep6-10  6.070e-01  3.551e-01   1.709  0.10283    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for quasipoisson family taken to be 1.355392)

    Null deviance: 1473.124  on 49  degrees of freedom
Residual deviance:   27.468  on 20  degrees of freedom
AIC: NA

Number of Fisher Scoring iterations: 16
\end{Soutput}
\end{Schunk}

\begin{Schunk}
\begin{Sinput}
> anova(model2,test="F")
\end{Sinput}
\begin{Soutput}
Analysis of Deviance Table

                Df Deviance Resid. Df Resid. Dev        F    Pr(>F)    
NULL                               49    1473.12                       
Age              4  1297.52        45     175.60 239.3263 < 2.2e-16 ***
Ethnicity        4    20.21        41     155.39   3.7285  0.020056 *  
NZDep            1     8.62        40     146.77   6.3573  0.020285 *  
Age:Ethnicity   16    94.49        24      52.28   4.3569  0.001223 ** 
Ethnicity:NZDep  4    24.82        20      27.47   4.5771  0.008694 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

The results are:
\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-024}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-025}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-026}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-028}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-029}
\end{center}
\caption{test}
\label{fig:test}
\end{figure}

\section{Regression analyses results}
\label{sub:regression_results}
The results are:

@@

Deviance Residuals: 
     Min        1Q    Median        3Q       Max  
-2.76316  -0.52409   0.04736   0.49883   1.92918

Coefficients:
                          Estimate Std. Error t value Pr(>|t|)    
(Intercept)               -6.46738    0.11483 -56.324  < 2e-16 ***
Age3-5                    -0.91290    0.20539  -4.445 0.000278 ***
Age6-17                   -0.76175    0.13427  -5.673 1.81e-05 ***
Age18-24                  -1.50501    0.19366  -7.772 2.57e-07 ***
Age25+                    -2.95073    0.16079 -18.352 1.51e-13 ***
EthnicityAsian             0.05304    0.32498   0.163 0.872072    
EthnicityMaori             0.21245    0.19935   1.066 0.299919    
EthnicityPacific           1.16032    0.20429   5.680 1.78e-05 ***
NZDep6-10                 -0.21192    0.08548  -2.479 0.022713 *  
Age3-5:EthnicityAsian     -2.03154    1.38265  -1.469 0.158112    
Age6-17:EthnicityAsian    -1.00742    0.47171  -2.136 0.045940 *  
Age18-24:EthnicityAsian   -0.83115    0.59409  -1.399 0.177917    
Age25+:EthnicityAsian      0.42108    0.44334   0.950 0.354143    
Age3-5:EthnicityMaori     -0.38639    0.40959  -0.943 0.357344    
Age6-17:EthnicityMaori    -0.08552    0.24685  -0.346 0.732812    
Age18-24:EthnicityMaori   -0.58278    0.44835  -1.300 0.209209    
Age25+:EthnicityMaori     -1.04821    0.52418  -2.000 0.060036 .  
Age3-5:EthnicityPacific   -2.13163    0.81388  -2.619 0.016882 *  
Age6-17:EthnicityPacific  -1.45411    0.33466  -4.345 0.000349 ***
Age18-24:EthnicityPacific -0.98266    0.51289  -1.916 0.070546 .  
Age25+:EthnicityPacific   -0.61270    0.43665  -1.403 0.176692    
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

(Dispersion parameter for quasipoisson family taken to be 1.776063)

    Null deviance: 1453.62  on 39  degrees of freedom
Residual deviance:   36.83  on 19  degrees of freedom
AIC: NA

Number of Fisher Scoring iterations: 5
\end{Soutput}
\end{Schunk}

@@

              Df Deviance Resid. Df Resid. Dev        F   Pr(>F)    
NULL                             39    1453.62                      
Age            4  1304.20        35     149.43 183.5795 6.73e-15 ***
Ethnicity      3    20.00        32     129.43   3.7529 0.028500 *  
NZDep          1    10.70        31     118.74   6.0236 0.023932 *  
Age:Ethnicity 12    81.91        19      36.83   3.8430 0.004459 ** 
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{Soutput}
\end{Schunk}

And plotted in (Figure~\ref{fig:regn}).
\begin{figure}
     \centering
     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
     \caption{Top left: distribution of the cases per category; Top right: Cases per Category; Bottom left: Model predictions (x) versus case data (y); and Bottom right: the residuals of the model}
     \label{fig:regn}
\end{figure}

The results of the model suggest that, apart from over representation of MELAA, older people over 25 years of age are significantly less likely to be infected with measles than the very young (0--2 years old), but Pacific peoples are at greater risk than Europeans and others (as can also be seen in Figure 1), but not the 6--17 years among them.

In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected. The very young are still most likely to be infections, but older teenagers are more likely to be repreested (Figure N). This pattern suggests that imporving vaccination coverage in the young is reducing the burden of measles in those age categories. See discussion regarding vaccination protocols.

Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data. We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g. meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes. Thus the data gap that we have that will hinder us providing fine scale risk maps is:
\begin{itemize}
\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
\end {itemize}
An additional data set that would enable us to develop the understanding of measles importation risk is:
\begin {itemize}
\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
\end{itemize}

@@

To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions and the simple relationship between the proportion requiring vaccination, $p_c$, and $R_0$. Specifically, we use the simple relationship:

\begin{equation} \label{eq:pc}
p_c = 1-(1/R0)
  \end{equation}

We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population, estimated from the national level vaccination coverage data provided by the Ministry of Health for this work. Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one.

And discuss the implications for this simplification in the discussion. However, briefly, there are several caveats to our use of this equation:
\begin {itemize}
\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
\end{itemize}

discussion of simplifications and use of other av:
\begin {itemize}
\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
\end{itemize}

We then used the proportions of the population that were estimated be require vaccination as estimated above, and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.

\subsection{Modelling results}

@@

The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\ref{fig:hosp}).

\begin{figure}[h!]
\begin{center}
\includegraphics{interimreport2-039}
\end{center}
\caption{Number of cases attending hospital and stay duration}
\label{fig:hosp}
\end{figure}

Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital. Only eight cases spent a week or more in the hospital. Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days. The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
